---
title: "XAI3"
author: "Marc Vicedo Carboneres"
date: "24/5/2024"
output: html_document
---

```{r carga librerias necesarias}

library(dplyr)
library(plotly)
library(reshape2)
library(lubridate)
library(randomForestSRC)

```

EXERCISE:
Apply PDP to the regression example of predicting bike rentals. Fit a random 
forest approximation for the prediction of bike rentals (cnt). Use the partial 
dependence plot to visualize the relationships the model learned. Use the slides 
shown in class as model.

QUESTION:
Analyse the influence of days since 2011, temperature, humidity and wind 
speed on the predicted bike counts.

```{r carga y transformación de datos para aplicar el modelo de random forest}

days <- read.csv("day.csv")
hour <- read.csv("hour.csv")

days$dteday <- as_date(days$dteday)
days_since <- select(days, workingday, holiday, temp, hum, windspeed, cnt)
days_since$days_since_2011 <- int_length(interval(ymd("2011-01-01"), days$dteday)) / (3600*24)
days_since$SUMMER <- ifelse(days$season == 3, 1, 0)
days_since$FALL <- ifelse(days$season == 4, 1, 0)
days_since$WINTER <- ifelse(days$season == 1, 1, 0)
days_since$MISTY <- ifelse(days$weathersit == 2, 1, 0)
days_since$RAIN <- ifelse(days$weathersit == 3 | days$weathersit == 4, 1, 0)
days_since$temp <- days_since$temp * 47 - 8
days_since$hum <- days_since$hum * 100
days_since$windspeed <- days_since$windspeed * 67

```


```{r we create the rf model and analyse the influence of days since 2011, temperature, humidity and wind speed on the predicted bike counts.}
rf_model <- rfsrc(cnt~., data=days_since)

output <- select(days_since, days_since_2011, temp, hum, windspeed, cnt)
num_rows <- nrow(days_since)
for(column in names(output)[1:4])
{
  for(index in 1:num_rows){
    temp_data <- days_since
    temp_data[[column]] <- days_since[[column]][index]
    prediction <- predict(rf_model, temp_data)$predicted
    output[[column]][index] <- sum(prediction) / num_rows
  }
}

```

EXERCISE:
Generate a 2D Partial Dependency Plot with humidity and temperature to 
predict the number of bikes rented depending of those parameters.

BE CAREFUL: due to the size, extract a set of random samples from the BBDD 
before generating the the data for the Partial Dependency Plot.

Show the density distribution of both input features with the 2D plot as 
shown in the class slides.

TIP: Use geom_tile() to generate the 2D plot. Set width and height to 
avoid holes.

```{r generate graphs}

plot1 <- ggplot(days_since, aes(x = temp, y = output$temp)) + geom_line() + ylim(c(0,6000)) +
  geom_rug(alpha = 0.1, sides = "b") + xlab("Temperature")

plot2 <- ggplot(days_since, aes(x = hum, y = output$hum)) + geom_line() + ylim(c(0,6000)) +
  geom_rug(alpha = 0.1, sides = "b") + xlab("Humidity")


# Usar subplot de plotly para organizar los gráficos
subplot(plot1, plot2, shareY = TRUE, shareX = FALSE, titleX = TRUE)

```